---
title: "Coin tossing simulation investigation"
date: "Saturday, January 03, 2015"
author: "Eamonn O'Brien"
output:
  pdf_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 8
    number_sections: yes
  html_document: default
  word_document: default
---


```{r set-options, echo=FALSE, cache=FALSE, warning = FALSE}

        rm(list=ls())

        set.seed(123)
        startTime <- proc.time()
        library(knitr)
        options(width=120)
        opts_chunk$set(comment = "", warning = FALSE, message = FALSE,
                       echo = TRUE, tidy = FALSE, size="tiny",  cache=TRUE,
                       progress=TRUE,
                       cache.path = 'program_Cache/',
                       fig.path='figure/')
         
        knitr::knit_hooks$set(inline = function(x) {
          knitr:::format_sci(x, 'md')
        })
```

```{r prerequites, echo=FALSE}
wd <- getwd()
setwd(wd)
opts_knit$set(root.dir = wd)

```
**This note from David Spiegelhalter inspired me to look at coin tossing problems : 'What are the chances of 4 or more heads or 4 or more tails in _n_ coin tosses (with a fair coin)?...about 47% chance from 10 flips...with 11 ~ 51%...'**

# What’s the probability of x consecutive results with a limited number of tosses?

## Create a function

```{r function}
tmpfun <- function(x, y) {{
  
  foo <- rbinom(x, 1, 1/2)  # binomial distribution of fair coin tosses
  rx <- rle(foo)            # function to examine runs in a sequence
  
   a <- any( rx$lengths  >= y )                       # any run of 0s or 1s in sequence?
  
   b <- any( rx$lengths[ rx$values %in% c(0) ] >= y ) # only run of 0s in sequence?
  
   c <- any( rx$lengths[ rx$values %in% c(1) ] >= y ) # only run of 1s in sequence?
  
   d <- any(rx$lengths[ rx$values %in% c(0)]>= y) &   # both 0s and 1s in a sequence?
   any(rx$lengths[ rx$values %in% c(1)]>= y)

   f <- (all(rx$lengths[ rx$values %in% c(0)] < y) & any(rx$lengths[ rx$values %in% c(1)] >= y))

   g <- any(rx$lengths[ rx$values %in% c(0)] >= y)& all(rx$lengths[ rx$values %in% c(1)] < y )  
}
ret = list()  # think of a venn diagram
ret$a  = a # any run of desired length , either heads or tails or both
ret$b  = b # any run of desired length , tails only
ret$c  = c # any run of desired length , heads only
ret$d  = d # any run of desired length , of both heads and tails
ret$f  = f # any run of desired length , with a (single) run of heads and so no run of tails
ret$g  = g # any run of desired length , with a (single) run of tails and so no run of heads

return(ret)

}
```
\newpage
## Execute function and manage the output
```{r exec function}
set.seed(123)   # reproducible result
run <- 4        # number of consecutive tosses
tosses <- 10    # total number of tosses

z <- replicate(1e04 , tmpfun(tosses, run))  # execute function large number of times
new_mat  <- array(as.numeric(z), dim(z))    # manage the output
foo <- as.data.frame(as.matrix(t(new_mat))) # create a data frame
head(foo)                                   # look at first 5 rows of data
res <- apply(foo,2,mean)                    # calculate column averages and examine 
```
## The probability of `r run` runs or more in `r tosses` tosses is `r res[1]`. So `r 1-res[1]` of the time no run of `r run` heads and/or tails observed.

```{r plot1, echo=FALSE }
library(VennDiagram)
grid.newpage()
venn.plot <- draw.pairwise.venn(area1 = res[6]+res[4], area2 = res[5]+res[4], cross.area = res[4], category = c("tails",   "heads"), 
        fill = c("blue", "red"), 
        lty = "blank")
grid.draw(venn.plot)
```


```{r exec function2, echo=FALSE}
run <- 4          # number of consecutive tosses
tosses <- 11      # total number of tosses

z <- replicate(1e04 , tmpfun(tosses, run))    # execute function large number of times
new_mat  <- array(as.numeric(z), dim(z))    # manage the output
foo <- as.data.frame(as.matrix(t(new_mat))) # create a data frame
res <- apply(foo,2,mean)                      # calculate column averages and examine
```
## The probability of `r run` runs or more in `r tosses` tosses is `r res[1]`. So `r 1-res[1]` of the time no run of `r run` heads and/or tails observed.

```{r plot2, echo=FALSE}
grid.newpage()
grid.newpage()
venn.plot <- draw.pairwise.venn(area1 = res[6]+res[4], area2 = res[5]+res[4], cross.area = res[4], category = c("tails",   "heads"), 
        fill = c("blue", "red"), 
        lty = "blank")
grid.draw(venn.plot)
```

```{r exec function3, echo=FALSE}
run <- 4         # number of consecutive tosses
tosses <- 15     # total number of tosses

z <- replicate(1e04 , tmpfun(tosses, run))    # execute function large number of times
new_mat  <- array(as.numeric(z), dim(z))    # manage the output
foo <- as.data.frame(as.matrix(t(new_mat))) # create a data frame
res <- apply(foo,2,mean)                      # calculate averages
```
## The probability of `r run` runs or more in `r tosses` tosses is `r res[1]`.  So `r 1-res[1]` of the time no run of `r run` heads and/or tails observed.

```{r plot3, echo=FALSE}
grid.newpage()
grid.newpage()
venn.plot <- draw.pairwise.venn(area1 = res[6]+res[4], area2 = res[5]+res[4], cross.area = res[4], category = c("tails",   "heads"), 
        fill = c("blue", "red"), 
        lty = "blank")
grid.draw(venn.plot)
```

\newpage 
# Examine how many tosses are required for _x_ consecutive results in a row? We are not concerned if it is heads or tails...whichever consecutive result comes first.

## Create a function
```{r function2}
coin <- c(0,1)

ComputeNbTosses <- function(targetTosses) {
  nbTosses <- 0 
  nbHeadsInRow <-  0
  nbTailsInRow <-  0
  allTosses <-  c()
  
  # keep tossing unless we reach target for either heads or tails
  while  (nbHeadsInRow < targetTosses & nbTailsInRow < targetTosses)  {
    
    toss = sample(coin,1,T)          # toss an unbiased coin
    allTosses = c(allTosses, toss)   # accumulate the tosses
    
    # count occurrences of runs of heads and of tails
    if (toss == 1) {nbHeadsInRow = nbHeadsInRow + 1} else {nbHeadsInRow = 0}
    if (toss == 0) {nbTailsInRow = nbTailsInRow + 1} else {nbTailsInRow = 0}
    
    nbTosses = nbTosses + 1         # count the tosses
  }

  ret = list()
  ret$nbTosses = nbTosses           # record the count of the tosses
  # ret$allTosses = allTosses       # collect this if you want to check simulation works as expected
  return(ret)
  
} 
```
\newpage
## Execute function for one scenario and manage the output
```{r exec function no 2}
set.seed(4321)                               # reproducible result
n <- 4                                         # number of consecutive tosses
result <- replicate(10000, ComputeNbTosses(n)) # execute function large number of times
```
## Summary of results
```{r summarise }
summary(unlist(result))                     
quantile( unlist(result),c(.001, .025,.1,.9,.95,.975,.99, .999))
```
**Therefore on average (mean) `r mean(unlist(result))` tosses are required to obtain `r n` consecutive results. It is unlikely (less than 5% of the time) that `r n` runs will require more than `r quantile( unlist(result),c(.95))` tosses.**



## Perform the simulation en masse (simulate for a number of scenarios in one function)
```{r create array for data}
set.seed(123)   ##for a reproducible result
I <-  c(2:10)     ##consecutive runs of interest
nrep <- 1000    ##simulations to run on each scenario

pwpr <- array(dim=c(length(I),nrep,1),
              dimnames=list(consecutive=I , 
                            simulation=seq(nrep),
                            Estimate=c("mean"))
)
```
## Loop
```{r loop through the function}
for (i in seq_along(I)) {
      pwpr[i,,]  <- plyr::raply(nrep, 
                     unlist(ComputeNbTosses(I[i]))[1][[1]]
                        )
      
    }  
```
\newpage
## Summary
```{r manage the output}
pwpr[,1:10,] # first 10 simulations  
p0 <- function(x) {formatC(x, format="f", digits=0)}
```
## Means, rownames are the desired run of consecutive tosses and below the average number of tosses
```{r manage the output2}
(resmeans  <-  (apply(pwpr,c(1),mean,na.rm=TRUE) ))
```
## Percentiles
```{r manage the output3}
(resci <-  (apply(pwpr,c(1 ),
                  quantile,c(.001, .025,.1,.25,.5,.75,.9,.95,.975,.99, .999), na.rm=TRUE)))
``` 

```{r sample a column at random, cache=FALSE, echo=FALSE, results='hide'}
x=sample(I,1)
```
**Interpret one example: The median number of tosses required  for `r x` consecutive runs of a fair coin are `r resci[5,x-1]`. It is unlikely (less than 5% of the time) that to observe `r x` consecutive runs will require more than `r resci[8,x-1]` tosses.**


# References   

http://stackoverflow.com/questions/21392665/homework-simulating-coin-tosses-until-consecutive-heads-using-r  
https://twitter.com/d_spiegel/status/544422589670916096   
http://stats.stackexchange.com/questions/21825/probability-over-multiple-blocks-of-events  
http://r.789695.n4.nabble.com/Plot-does-not-show-in-R-td4693637.html
http://math.stackexchange.com/questions/364038/expected-number-of-coin-tosses-to-get-five-consecutive-heads  .5^2
http://www.cs.cornell.edu/~ginsparg/physics/INFO295/mh.pdf  
error on bottom of page 5, n=5 not reported  

(2^(n+1))-2

2*(2^n-1)

http://stats.stackexchange.com/questions/91518/waiting-time-for-successive-occurrences-of-a-result-when-rolling-a-die?rq=1  
p<-.5^n
2*(1-p)/p

http://stats.stackexchange.com/questions/12174/time-taken-to-hit-a-pattern-of-heads-and-tails-in-a-series-of-coin-tosses  
http://stats.stackexchange.com/questions/91518/waiting-time-for-successive-occurrences-of-a-result-when-rolling-a-die?rq=1  
http://stats.stackexchange.com/questions/126884/how-many-times-do-i-have-to-roll-a-die-to-get-six-six-times-in-a-row  
http://math.stackexchange.com/questions/192177/how-many-times-to-roll-a-die-before-getting-two-consecutive-sixes

 
# Computing Environment

```{r, echo=FALSE}
#opts_knit$set(root.dir = wd)   ##THIS SETS YOUR WORKING DIRECTORY
options(width=70)
sessionInfo()
print(wd)
```
```{r echo=FALSE}
stopTime<-proc.time()
```
This took `r (stopTime-startTime)[1][[1]]` seconds to execute. 
