---
title: Perform contrasts in linear regression and survival analysis. Then moving on to examples of some KM plots
author: "Eamonn O'Brien"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 8
    number_sections: yes
  html_document: default
---

# Introduction

 
Showing how to perform contrasts in linear regression and survival analysis. Then moving on to examples of some KM plots


```{r set-options, echo=FALSE, cache=FALSE, warning = FALSE}

        rm(list=ls())
        set.seed(874)
        startTime<-proc.time()
        library(knitr)
        options(width=120)
        opts_chunk$set(comment = "", warning = FALSE, message = FALSE,
                       echo = TRUE, tidy = FALSE, size="tiny",  cache=FALSE,
                       progress=TRUE,
                       cache.path = 'program_Cache/',
                       fig.path='figure/')
         
        knitr::knit_hooks$set(inline = function(x) {
          knitr:::format_sci(x, 'md')
        })
        
        
         
```

```{r prerequisites, echo=FALSE ,eval=TRUE}
 
        where<-"home" #this is used in the sourced program 
 
        path <- "GIT\\programs"  
     
        work<-    paste("X:/", path, sep = "")
        nonwork<- paste("~/", path, sep = "")
        if (where=="home") {wd<- nonwork} else {wd<-work}
        
        path2 <- " "  
        work2<-    paste("X:/", path2, sep = "")
        nonwork2<- paste("~/X/", path2, sep = "")
        
        if (where=="home") {wd2<- nonwork2} else {wd2<-work2}
        
        work3<-    paste("X:/FUNCTIONS/R", sep = "")
        nonwork3<- paste("~/X/FUNCTIONS/R", sep = "")
        
        if (where=="home") {wd3<- nonwork3} else {wd3<-work3}
        setwd(wd)
        opts_knit$set(root.dir = wd)      ##THIS SETS YOUR WORKING DIRECTORY
        
 
```
 
 
```{r preliminaries perhaps , echo=FALSE, results='hide'}


        list.of.packages <- c("rms","ggplot2", "reshape", "rethinking")
        
        new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
        if(length(new.packages)) install.packages(new.packages)
        
        sapply(X = list.of.packages, require, character.only = TRUE)


        p3 <- function(x) {formatC(x, format="f", digits=3)}
        p4 <- function(x) {formatC(x, format="f", digits=4)}
        p2 <- function(x) {formatC(x, format="f", digits=2)}
         p1 <- function(x) {formatC(x, format="f", digits=0)}
        # p1 <- function(x) {print(formatC(x, format="f", digits=1),quote=FALSE)}
        # p2 <- function(x) {print(formatC(x, format="f", digits=2),quote=FALSE)}
        # p3 <- function(x) {print(formatC(x, format="f", digits=3),quote=FALSE)}
        # p4 <- function(x) {print(formatC(x, format="f", digits=4),quote=FALSE)}
        #perhaps help colour plot text based on loop count
        is.even <- function(x){ x %% 2 == 0 }
  
```            

# linear regression and contrasts (t dist)

```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE }  

       n <- 200
       intercept <- 100
       noise <- 5   
       # Hyp the effect of age on outcome, a unit change in outcome over the whole age range
       # so ~0.02 effect on outcome for each increase in age of 1 year
       effect.of.age <- 1/(65-18)       
       effect.of.sex <- 5         # Hypothesize the effect of sex on outcome
       effect.of.treatment <- 1   # Hypothesize the treatment effect on outcome
       
       # Hypothesize baseline version of response on outcome, 
       # a unit increase results in 0.5 increase in outcome, 
       # strong predictor so 5 over the baseline range
       effect.of.baseline <- 1/2          
                                      # random error 
   
       # covariate effects
       baseline <- runif(n,100,110)                 # baseline effect
       treat <- 1*(runif(n)<0.5)                    # randomised treatment effect
       fact <- c(rep("a", n/4), rep("b", n/4), rep("c", n/4), rep("d", n/4))  # Generate grp
       age <- sample(x=18:65, size=n, replace=TRUE) # Generate age covariate
    
      # create the outcome, with random error
        Y <- 2*(fact=="b")+  3*(fact=="c")+  4*(fact=="d") +
          effect.of.age*age + effect.of.baseline*baseline +
          effect.of.treatment*treat + rnorm(n, intercept, noise)
    
      # prepare for analysis
       d <- data.frame(Y=Y, baseline=baseline, treat=treat, 
                      fact=factor(fact),  age=age)
      
       dd <- datadist(d, data=d)
       options(datadist="dd")
    
```            

# my function for contrasts (t dist)

```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE }         
       
     lincom <- function (object, ref, comp) {
        
        var <- vcov(object, regcoef.only = TRUE, intercepts = "none")

        c1<-var[, grepl(comp, colnames(var))  ]
        var1<-c1[ grepl(comp, names(c1))  ]
        
        c1<-var[, grepl(ref, colnames(var))  ]
        var2<-c1[ grepl(ref, names(c1))  ]
        
        c1<-var[, grepl(comp, colnames(var))  ]
        cov<-c1[ grepl(ref, names(c1))  ]
        
        
        comp1<-object$coef[grepl(comp,   names(object$coef ))   ][[1]]
        ref1<- object$coef[grepl(ref,    names(object$coef ))   ][[1]]
        
        df <- qt(.975,  object$df.residual) # from harrell contrast.rms
        
        namez<-c("comp v", "ref", "est", "Lower95%CI", "Upper95%CI")
        res<-c(comp,ref,(comp1-ref1),
                (comp1-ref1+c(-1,1)*
                            df*sqrt(var1[[1]]+var2[[1]]-2*cov[[1]])))
                
               res <- c( res[1:2], sprintf(fmt="%.15s", res[3:5]))
  
        names(res)<-namez
        return(res)
        
}
      
```            

# variance covariance matrix


```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE } 

     f <- ols(Y ~ baseline + fact + age + treat, d)
     print(kable(vcov(f)))
     
```            

\clearpage

# confidence intervals

```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE } 

 
    d$fact <- relevel( d$fact, ref="a")           # no need for this first time
    f <- ols(Y ~ baseline + fact + age + treat, d)
    print(kable(confint(f)))
    
    
```            

# my function, compare level c to level d


```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE }     

    print(kable (lincom(f, ref="fact=d", comp= "fact=c") ) )
    
```            

# relevel to get a contrast comparing level c to level d


```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE } 

    d$fact <- relevel( d$fact, ref="d")
    f <- ols(Y ~ baseline + fact + age + treat, d)
    print(kable(confint(f)    )) 
     
```            

# Harrell's contrast function

```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='markup' , cache=TRUE } 

    x <- contrast(f, list(fact="d"), list(fact="c")  )
    print(x, latex=TRUE, file='') 
     
```            

\clearpage

# survival analysis and contrasts (z dist)

```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE }  

  
        n <- 1000
        set.seed(731)
        age <- 50 + 12*rnorm(n)
        label(age) <- "Age"
        grp <- factor(sample(c('a','b','c','d'), n, 
                             rep=TRUE, prob=c(.25, .25,.25,.25)))
        cens <- 15*runif(n)
        
        h <- .02*exp(.04*(age-50)+.4*(grp=='b')+.6*(grp=='c')+.8*(grp=='d'))
        
        dt <- -log(runif(n))/h
        label(dt) <- 'Follow-up Time'
        
        e <- ifelse(dt <= cens,1,0)
        dt <- pmin(dt, cens)
        units(dt) <- "Year"
        dd <- datadist(age, grp)
        options(datadist='dd')
        
        S <- Surv(dt,e)
        f <- cph(S ~ (age) + grp, x=TRUE, y=TRUE)
        # cox.zph(f, "rank")             # tests of PH
        print(kable(anova(f)))
        #print(kable(f))

```            

# variance covariance matrix


```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE } 

       print(kable(vcov(f)))
     
```        

\clearpage

# Another function for survival analysis contrasts              
                
```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE } 
 

      lincom.cph <- function (object, ref, comp) {
        
        var <- vcov(object, regcoef.only = TRUE, intercepts = "none")
        
        c1<-var[, grepl(comp, colnames(var))  ]
        var1<-c1[ grepl(comp, names(c1))  ]
        
        c1<-var[, grepl(ref, colnames(var))  ]
        var2<-c1[ grepl(ref, names(c1))  ]
        
        c1<-var[, grepl(comp, colnames(var))  ]
        cov<-c1[ grepl(ref, names(c1))  ]
        
        
        comp1<-object$coef[grepl(comp,   names(object$coef ))   ][[1]]
        ref1<- object$coef[grepl(ref,    names(object$coef ))   ][[1]]
        
        namez<-c("comp v", "ref", "HR", "Lower95%CI", "Upper95%CI")
        res<-c (comp,ref,exp(comp1-ref1),
                exp(comp1-ref1+c(-1,1)*
                      1.96*sqrt(var1[[1]]+var2[[1]]-2*cov[[1]])) )
                
        res <- c( res[1:2], sprintf(fmt="%.5s", res[3:5]))
                
        names(res)<-namez
        return(res)
        
      }
      
```            

# my function, compare level c to level d


```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE }
       
      print(kable(lincom.cph(f, ref="grp=d", comp="grp=c")))
      
```            

# Harrell function, compare level c to level d


```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE }
      print(kable(summary(f, grp=c('d'), est.all=F) ))  
      
```        

# Harrell again
                
                
```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='markup' , cache=TRUE }     

      x <- contrast(f, list(grp="c"), list(grp="d")  )
      print(x, X=F, fun=exp  )
```        
                
                
```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE } 
 
      plot( summary(f, grp=c('a'), est.all=F), log=T )

```        

\clearpage

# Explore simple survival plots              
                
```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE } 

      library(ggfortify)
      library(survival)
      f <- survfit(S ~   grp )
      autoplot(f)
      autoplot(f, facets = TRUE, nrow=4)

```        

\clearpage

# Explore simple survival plots              
                
```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE } 

      library("survminer")
      ggsurvplot(f, pval = TRUE, 
                 break.time.by = 3,
                 risk.table = TRUE,
                 risk.table.col = "strata",
                 risk.table.height = 0.5, 
                 ggtheme = theme_bw(),
                 legend.labs = c("A", "B", "C", "D"))

```        

\clearpage

# Explore simple survival plots              
                
```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE } 
 
      f2 <- npsurv(S ~ grp)
       
      x <- c("#e41a1c","#377eb8","#4daf4a","#984ea3")
      
       #x <- rainbow(40) 
       #x <- x[c(1,5,36,27)]
       
      
      survplot(f2,  n.risk=TRUE, levels.only=T, conf.int=T,
               aehaz=TRUE,
               conf=c("bands"), 
               col.fill=gray(seq(.95, .75, length=4)),
               #col.fill= c(rgb(0,0,0,0.1)), 
               type=c("kaplan-meier"),
               lty=1,  col=x, xlab="Months", abbrev.label=T, 
               label.curves = list(keys = "lines"), #bty='n',
               y.n.risk= 0, cex.n.risk=.6, time.inc=2)
      #lines(f2,  col= x) 

```        

\clearpage

# Explore simple survival plots              
                
```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE } 

        x <- rainbow(40) 
        x <- x[c(1,5,36,27)]
        
        survplot(f2,  n.risk=TRUE, levels.only=T, conf.int=T,
                 aehaz=TRUE,
                 conf=c("bands"), 
                 col.fill=gray(seq(.94, .88, length=4)),
                 #col.fill= c(rgb(0,0,0,0.1)), 
                 type=c("kaplan-meier"),
                 lty=4,  col=x,
                 xlab="Months", abbrev.label=T, 
                 label.curves=list(method="arrow", cex=.8),
                 #label.curves=list(keys="lines"), 
                 y.n.risk= 0, cex.n.risk=.5, time.inc=2)

```        

\clearpage

# Explore simple survival plots              
                
```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='asis' , cache=TRUE } 

      setwd(wd3)
      source("Alternative KM plot.R", echo = F) # note this code is stored elsewhere
      setwd(wd)


kmplot(f2, mark='¬', simple=F,
  xaxis.at=seq(0,14,2), 
  xaxis.lab=seq(0,14,2), # n.risk.at
  lty.surv=c(1,1), lwd.surv=1, col.surv=rainbow(4), # survival.curves
       lty.ci=0, lwd.ci=.7, col.ci=rainbow(4), #0 ci not plottrd
   group.names=c('a','b','c','d'),
	group.order=c(1,2,3,4), # order of appearance in the n.risk.at table and legend.
	extra.left.margin=6, label.n.at.risk=T, draw.lines=TRUE,
	cex.axis=0.8, xlab='Time to event (months)', ylab='Survival probability', # labels
	grid=TRUE, lty.grid=1, lwd.grid=1, col.grid=grey(.9),
	legend=T, loc.legend='bottomleft',
	cex.lab=1.1, xaxs='r', bty='L', las=1, tcl=-.2  # other parameters passed to plot()
)
title(main='test', adj=.1, font.main=1, line=0.5, cex.main=1)

```        
 
```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='markup' , cache=TRUE } 
 # exponential distribution hazard rate estimates
print(f2$numevents/f2$exposure, digits=4)


```

\clearpage
\pagebreak 


# Computing Environment

```{r, echo=FALSE}
#opts_knit$set(root.dir = wd)   ##THIS SETS YOUR WORKING DIRECTORY
options(width=70)
sessionInfo()
#print(wd)
```
```{r echo=FALSE}
stopTime<-proc.time()
```
This took `r (stopTime-startTime)[1][[1]]` seconds to execute. 
  
