---
title: "Pierre Rémond de Montmort's matching problem 1708"
date: "Sunday, May 21, 2016"
output: html_document
---

 
```{r set-options, echo=FALSE, cache=FALSE  , message=F, warning=F} 
    rm(list=ls()) 
    startTime<-proc.time()
    options(width=120)
    require(knitr)
    knitr::opts_chunk$set(comment = "", warning = FALSE, message = FALSE, 
                   echo = TRUE, tidy = FALSE, size="tiny",  
                   cache=TRUE, progress=TRUE, cache.path = 'Coin_Cache/', 
                   fig.path='figure/') 

``` 
************************

**The biggest coincidence of all would be if there were no coincidences.**

**`In its simplest form the problem of coincidences (matches, recontres) may be formulated as follows: Let there be n objects numbered from 1 to n, and let them be ordered at random, assuming that the n! permutations are equally probable. A coincidence occurs if object number i is found at the ith place. The problem is to find the number of permutations with at least one coincidence or, equivalently, the probability of at least one coincidence.**

**When the sample size is 5, what's the probability of observing at least one match? For n = 5, the argument is as follows: The number of permutations of 5 cards is 5! = 120. Among these there are 24 in which 1 is in first place, 18 in which 2 is in second place without 1 being first, 14 in which 3 is in third place without 1 being first or 2 being second, 11 in which 4 is in fourth place without 1 being first, 2 being second or 3 being third, and, finally, 9 in which 5 is in fifth place, the other four being out of their places. The probability of at least one coincidence is therefore:'** [A. Hald]

```{r  , echo=TRUE, cache=FALSE  , message=F, warning=F} 

   (numerator <- 24 + 18 + 14 + 11 + 9)
   (denominator <- 5 * 4 * 3 * 2 * 1)        
    numerator / denominator
 
``` 

************************

**Plot a simulated example, do we observe at least one match?**


```{r  , echo=TRUE, cache=FALSE  , message=F, warning=F} 

    set.seed(nchar("de Montmort"))       # Reproducible results
    z <- round(runif(1, 5, 1000), 0)     # Randomly choose a sample size
    f <- sample(1:z, z)                  # Shuffle the deck
    y <- 1:z                             # The original order
    res <- (sum(f==y)) > 0               # Does at least one card match the original order?
    which(f==y)                          # When do we encounter the matches?
    
    # Plot the single simulation
    
    colr <- ifelse(f==y, 'red', 'black') 
    cexr <- ifelse(f==y, 2, 1)
    pchr <- ifelse(f==y, 16, 20)
    
    plot(f, col = colr, cex = cexr, pch = pchr,
    main = paste("N=", z, ", Do we see at least 1 match?", res, ", Matches found", sum(f==y), sep = " ")) 
    abline(0,1)


``` 

************************

**Simulate a large number of times**

```{r  , echo=TRUE, cache=FALSE  , message=F, warning=F} 

    # On average what is the chance of observing at least one match?
    
    z <- f <- y <- res <- NULL
    
    deM <- function() {
      
      z <- round(runif(1, 5, 1e4), 0)   # Let us vary the sample size each time
      f <- sample(1:z, z)               # Shuffle the deck
      y <- 1:z                          # The original order
      res <- sum(f==y) > 0              # Does at least one card match the original order?
      
    }
    
    mean(replicate( 1e5, deM() ))       # Execute function large number of times

```    
    
************************

**Simulate a large number of times with a sample size 100, another version**

```{r  , echo=TRUE, cache=FALSE  , message=F, warning=F}     
    
  n <- 1e6    
  samp <- 100   
  
  foo <- numeric( n )
    for( i in 1:n ) foo[i] <- sum( sample( 1:samp, samp ) == 1:samp ) ==  0
  1-mean( foo ) 
      
    
``` 

************************

**Analytical solution 1 - 1/e**

```{r  , echo=TRUE, cache=FALSE  , message=F, warning=F} 
 

    1 - ( 1 / exp(1) )  
    
    ( exp(1) - 1 ) / exp(1)  # same

    
``` 

************************

**Reference** 

A. Hald, A History of Probability and Statistics and their Applications Before 1750, Wiley, New York, 1990.  

**Computing Environment**

```{r, echo=FALSE}
sessionInfo()
 
```
```{r echo=FALSE}
stopTime<-proc.time()
```
This took `r (stopTime-startTime)[1][[1]]` seconds to execute.