---
title: "Nested Variance Components Analysis (Bayesian)"
author: "Eamonn O'Brien"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document: default
  pdf_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 8
    number_sections: yes
---

### Introduction

Return to this later


```{r set-options, echo=TRUE, cache=FALSE, warning = FALSE}

        rm(list=ls())
        set.seed(874)
        startTime<-proc.time()
        library(knitr)
        options(width=120)
        opts_chunk$set(comment = "", warning = FALSE, message = FALSE,
                       echo = TRUE, tidy = FALSE, size="tiny",  cache=FALSE,
                       progress=TRUE,
                       cache.path = 'program_Cache/',
                       fig.path='figure/')
         
        knitr::knit_hooks$set(inline = function(x) {
          knitr:::format_sci(x, 'md')
        })
```

```{r prerequisites, echo=TRUE}
 
        where<-"home" #this is used in the sourced program 
 
        path <- ""  
     
        work<-    paste("X:/", path, sep = "")
        nonwork<- paste("~/X/", path, sep = "")
        if (where=="home") {wd<- nonwork} else {wd<-work}
        
        path2 <- ""
        
        work2<-    paste("X:/", path2, sep = "")
        nonwork2<- paste("~/X/", path2, sep = "")
        
        if (where=="home") {wd2<- nonwork2} else {wd2<-work2}
        
        work3<-    paste("X:/FUNCTIONS/R", sep = "")
        nonwork3<- paste("~/X/FUNCTIONS/R", sep = "")
        
        if (where=="home") {wd3<- nonwork3} else {wd3<-work3}
        setwd(wd)
        opts_knit$set(root.dir = wd)  ##THIS SETS YOUR WORKING DIRECTORY
        
 
```
 
 
```{r preliminaries perhaps , echo=TRUE, results='hide'}


        list.of.packages <- c("lme4","rethinking", "rstanarm")
        
        new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
        if(length(new.packages)) install.packages(new.packages)
        
        sapply(X = list.of.packages, require, character.only = TRUE)


        p3 <- function(x) {formatC(x, format="f", digits=3)}
        p4 <- function(x) {formatC(x, format="f", digits=4)}
        p2 <- function(x) {formatC(x, format="f", digits=2)}
         p1 <- function(x) {formatC(x, format="f", digits=0)}
        # p1 <- function(x) {print(formatC(x, format="f", digits=1),quote=FALSE)}
        # p2 <- function(x) {print(formatC(x, format="f", digits=2),quote=FALSE)}
        # p3 <- function(x) {print(formatC(x, format="f", digits=3),quote=FALSE)}
        # p4 <- function(x) {print(formatC(x, format="f", digits=4),quote=FALSE)}
        #perhaps help colour plot text based on loop count
        is.even <- function(x){ x %% 2 == 0 }
  
```            

### Modelling

```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='markup' , cache=TRUE }  
 
   data(Glucose, package="VCA") # be specific, nlme also has a Glucose dataset, see 'data()'
    head(Glucose)
  
    dd <- Glucose
  
    dd$run <- rep(1:(length(dd$run)/2),each=2)

    # frequentist
    f <- lme4::lmer(result ~ (1 | day/run), data = dd, 
                  REML = TRUE, na.action = "na.omit")

    # Bayesian rethinking
    # bundle data and some indtructions  
    require(rethinking)
    
    data_list <- list( 
      y = dd$result, 
      day = dd$day, 
      run = rep(1:(length(dd$run)/2),each=2),
      start=list(sigma=sd(dd$result), 
                 sigma_day=sd(dd$result), 
                 sigma_run=sd(dd$result), 
                 a =mean(dd$result) ))
    
    
   # mux <- mean(dd$result)
    # model
    f2 <- alist(
      ( y ~ dnorm(mu,sigma)),
    
      mu <- a + bN[day] + bR[run] ,
      
      a ~ dnorm(244 ,10),
      bN[day] ~ dnorm(0, sigma_day ),
      bR[run] ~ dnorm(0, sigma_run ),
      sigma_run ~ dcauchy(0,1), 
      sigma_day ~ dcauchy(0,1), 
      sigma ~ dcauchy(0,1)
    )
 
    # execute
    rethink1 <- map2stan( f2 , 
                   data=data_list , 
                   iter=100000, warmup=2000, chains=8  )

 

    # Bayesian Rstan 
    require(rstanarm)
    ITER <- 500L
    CHAINS <- 2L
    CORES <- 1L
    SEED <- 12345

    rstan1 <- stan_lmer(result ~ (1 | day/run), data = dd, 
                        prior=cauchy(0,1), #default 2.5
                        prior_intercept = normal( 0,NULL),# default 10
                        prior_covariance = decov(shape =1, scale = 1),
                        adapt_delta = 0.999, chains = CHAINS, cores = 1,
                        seed = SEED)

```            

### Modelling

```{r echo=TRUE, eval=TRUE, tidy=FALSE, results='markup' , cache=TRUE }  
    
  print(f, digits=4)
   precis(rethink1, depth = 1, prob = 0.95, digits = 3)
   print(rstan1, digits=3)
    
    
    
    #     posterior_interval(test, pars = c("(Intercept)", "sigma" ))
# 
# 
#     stan_trac(test)
#     stan_hist(test)

    #launch_shinystan(post2)

#     y_rep <- posterior_predict(post2)
#     dim(y_rep)

      

# mat = as.matrix(test)
# 
# # get the random effect columns
# group_cols <- grep(" day:", colnames(mat), value = T)
# group_posteriors <- mat[, group_cols]
# sqrt(median(apply(group_posteriors,2, var)))
# 
# group_cols <- grep(":day:", colnames(mat), value = T)
# group_posteriors <- mat[, group_cols]
# sqrt(mean(apply(group_posteriors,1, var)))
# 
# group_cols <- grep("sigma", colnames(mat), value = T)
# group_posteriors <- mat[, group_cols]
# (median(group_posteriors))
# 
# 
# 
# 
# 
# ##########
# # extract posterior samples
# mat = as.matrix(test)
# 
# # get the random effect columns
# group_cols <- grep(":day", colnames(mat), value = T) #run
# group_posteriors <- mat[, group_cols]
# 
# # example variable slope term: b[var1_scaled donor:donor1]
# variable_names <- unique(gsub("b\\[(.*)\\s.*", "\\1", colnames(group_posteriors)))
# 
# # for each variable, compute its SD at each MCMC iteration
# var_group_summary <- sapply(variable_names, function(vname){
#   var_group_posteriors <- group_posteriors[, grep(vname, colnames(group_posteriors), value = T)]
#   group_sd <- sapply(1:nrow(var_group_posteriors), function(crow){
#     out <- sd(var_group_posteriors[crow, ])
#     return(out)
#   })
#   group_summary <- c(mean = mean(group_sd), quantile(x = group_sd, probs = c(0.5, 0.025, 0.975)))
#   return(group_summary)
# })
# return(var_group_summary)
# }
# ##########







# coef(test)

# cv <- loo(post2)
# 
# par(mfrow = 1:2, mar = c(5,3.8,1,0) + 0.1, las = 3)
# plot(cv, label_points = TRUE)




  
# set.seed(666)
# Omega <- rbind(
#   c(1, 0.3, 0.2),
#   c(0.3, 1, 0.1),
#   c(0.2, 0.1, 1)
# )
# sigma <- c(1, 2, 3)
# Sigma <- diag(sigma) %*% Omega %*% diag(sigma)
# N <- 100 
# y <- mvtnorm::rmvnorm(N, c(0,0,0), Sigma)
# 
# apply(y, 2, mean)
# apply(y, 2, sd)
 

    

```

### References 

https://groups.google.com/forum/#!topic/stan-users/F23kBsgqGcY
https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations
http://stla.github.io/stlapblog/posts/StanLKJprior.html
http://www.psychstatistics.com/2014/12/27/d-lkj-priors/
stan reference manual
https://cran.r-project.org/web/packages/rstanarm/vignettes/rstanarm.html # note-on-prior-beliefs-and-default-priors: The default priors in rstanarm are designed to be weakly informative, by which we mean that they avoid placing unwarranted prior weight on nonsensical parameter values and provide some regularization to avoid overfitting, but also do allow for extreme values if warranted by the data.
http://rstudio-pubs-static.s3.amazonaws.com/170731_a498f35a82f74e009b6d8def040e422f.html#extracting-posterior-samples
https://github.com/stan-dev/rstanarm/wiki/Prior-distributions  
http://datascienceplus.com/bayesian-regression-with-stan-part-1-normal-regression/

    
\clearpage
\pagebreak 


### Computing Environment



```{r, echo=FALSE}
#opts_knit$set(root.dir = wd)   ##THIS SETS YOUR WORKING DIRECTORY
sessionInfo()
print(wd)
```
```{r echo=FALSE}
stopTime<-proc.time()
```
This took `r (stopTime-startTime)[1][[1]]` seconds to execute. 
  

